---
title: "Prepare sample metadata and BMC library manifest"
author: "Mike"
date: today
format:
  html:
    fig-format: svg
---

This script reads the redacted sample-log exports and the BMC notification
email under `data-raw/`, joins them into two analysis-ready tables, and
writes them to `data/`:

- `data/processed-samples-metadata.rds`: one row per PR-experiment
  extraction; carries collection-side fields from the matching collected
  sample.
- `data/libraries.tsv`: one row per BMC sequencing library; carries
  all upstream fields via the (library → submission → aliquot → processed →
  collected) join chain.

```{r setup}
#| output: false
library(tidyverse)
library(fs)
library(here)
options(warn = 1)  # print warnings as they occur
```

## Read sample-log CSVs using the dictionary

```{r dict}
dict <- here("data-raw", "sample-log", "dictionary.csv") |>
  read_csv(show_col_types = FALSE) |>
  janitor::clean_names()
```

```{r import-function}
#' Read one sheet CSV, using the dictionary to set column types and rename.
read_log_sheet <- function(sheet_name, file) {
  d <- dict |> filter(sheet == sheet_name)
  col_types <- str_c(d$readr_col_type, collapse = "")
  x <- read_csv(file, col_types = col_types, na = c("", "NA", "#N/A", "none"))
  stopifnot(identical(names(x), d$column_name))
  rename_lookup <- set_names(d$column_name, d$new_column_name)
  x |>
    rename(all_of(rename_lookup)) |>
    filter(!if_all(everything(), is.na))
}
```

```{r read-sheets}
log_collected <- read_log_sheet(
  "Collected samples",
  here("data-raw", "sample-log", "collected-samples.csv")
) |>
  mutate(sample_received_time = hms::as_hms(sample_received_time))

log_processed <- read_log_sheet(
  "Processed samples",
  here("data-raw", "sample-log", "processed-samples.csv")
)

log_aliquots <- read_log_sheet(
  "Extraction aliquots",
  here("data-raw", "sample-log", "extraction-aliquots.csv")
)

log_sequencing <- read_log_sheet(
  "Submitted for sequencing",
  here("data-raw", "sample-log", "submitted-for-sequencing.csv")
)
```

## Parse BMC notification email

The BMC email lists one line per sequenced library in the format
`    D24-NNNNNN : <submission_sample_id> (<i7+i5>)`. We extract those
lines and build the library table.

```{r}
email_lines <- here("data-raw", "bmc-email.txt") |> read_lines()
lib_lines <- email_lines |>
  str_subset("^    D24-[0-9]+ : ")

provider <- tibble(raw = lib_lines) |>
  mutate(
    raw = raw |>
      str_replace("^ +", "") |>
      str_replace(" : ", " ") |>
      str_replace_all("\\(|\\)", "")
  ) |>
  separate(raw,
    into = c("library_id_short", "submission_sample_id", "index_sequences"),
    sep = " "
  ) |>
  mutate(
    library_id = str_c("240212Esv_6656R_", library_id_short),
    provider = "MIT BioMicro Center",
    provider_abbrev = "MIT BMC",
    # The first prep batch was D24-138001..138032; the second was
    # D24-138101 onward (a re-prep after a broken-tube incident).
    provider_library_number = str_extract(library_id, "(?<=D24-)[0-9]+$") |>
      as.integer(),
    provider_prep_batch = if_else(provider_library_number >= 138101, 2L, 1L),
    library_layout = "Paired-end",
    platform = "Illumina",
    instrument_model = "NovaSeq 6000"
  ) |>
  select(-provider_library_number, -library_id_short) |>
  relocate(library_id)
```

## Build BMC library manifest

Same `sample_source` / `sample_type` recodes apply here as for the
processed-samples table.

```{r}
bmc_libraries <- provider |>
  left_join(log_sequencing, by = "submission_sample_id") |>
  rename(extraction_aliquot_id = submission_extraction_aliquot_id) |>
  left_join(log_aliquots, by = "extraction_aliquot_id") |>
  left_join(log_processed, by = "processed_sample_id") |>
  left_join(log_collected, by = "collected_sample_id") |>
  mutate(
    sample_source = fct_recode(sample_source, "BOS" = "BST"),
    sample_type = fct_recode(sample_type, "airplane" = "triturator inflow")
  )

write_tsv(bmc_libraries, here("data", "libraries.tsv"))
```

## Build processed-sample metadata

Join processed samples to their collected-sample fields. Negative
controls have no collected sample, so we fill in their `sample_source`
/ `sample_type` / `sample_description` based on the `-NEG` token in
the processed-sample ID and pair them with the wastewater samples
they were processed alongside by setting their `collection_date` to
the processing date.

Two further recodes apply: `BST` → `BOS` for `sample_source`
(reflecting the publication name of the airport site), and
`triturator inflow` → `airplane` for `sample_type` (the descriptive
label used throughout the manuscript).

```{r}
processed_samples_metadata <- log_processed |>
  left_join(log_collected, by = "collected_sample_id") |>
  mutate(
    is_neg = str_detect(processed_sample_id, "-NEG"),
    sample_source = if_else(is_neg, "NEG", sample_source),
    sample_type = if_else(is_neg, "negative control", sample_type),
    sample_description = if_else(is_neg, "negative control", sample_description),
    collection_date = if_else(is_neg, processed_date, collection_date),
    sample_source = fct_recode(sample_source, "BOS" = "BST"),
    sample_type = fct_recode(sample_type, "airplane" = "triturator inflow") |>
      fct_relevel("sludge", "influent", "airplane")
  ) |>
  select(-is_neg)

write_rds(
  processed_samples_metadata,
  here("data", "processed-samples-metadata.rds")
)
```

## Quick sanity checks

```{r}
processed_samples_metadata |> count(sample_type)
processed_samples_metadata |> count(sample_source)
bmc_libraries |> count(submission_sample_id) |> count(n, name = "n_libs")
```
