---
title: "Prepare qPCR and Qubit data"
author: "Mike"
date: today
format:
  html:
    fig-format: svg
---

This script reads the raw QuantStudio Results CSVs under
`data-raw/qpcr/` and the Qubit measurements at
`data-raw/qubit/measurements.csv`, and writes two analysis-ready files
to `data/`:

- `data/qpcr.rds`: one row per qPCR well (SARS-CoV-2 / PMMoV targets).
- `data/qubit.rds`: one row per (sample × Qubit assay) measurement in
  long format, with concentration columns derived from the raw machine
  reading and the dilution factor.

```{r setup}
#| output: false
library(tidyverse)
library(fs)
library(here)
library(lubridate)
options(warn = 1)  # print warnings as they occur

source(here("code", "_helpers.R"))
```

## qPCR

```{r}
qpcr_files <- dir_ls(here("data-raw", "qpcr"), glob = "*_Results_*.csv") |>
  tibble(path = _) |>
  mutate(file = path_file(path))

qpcr_results <- qpcr_files |>
  mutate(data = map(path, read_qpcr_results_csv)) |>
  select(-path) |>
  unnest(data) |>
  mutate(
    # Fix the one plate where the NTC well was labeled differently
    task = if_else(sample == "NTC", "NTC", task),
    sample = if_else(sample == "NTC", NA_character_, sample),
    # Correct mislabeled sample IDs: some say "Neg" instead of "NEG", and some
    # carry a "PR-22####-" year typo that should read "PR-23####-".
    sample = str_replace(sample, "Neg", "NEG"),
    sample = str_replace(sample, "^PR-22", "PR-23")
  )

write_rds(qpcr_results, here("data", "qpcr.rds"))
```

## Qubit

The raw CSV has columns: `handler`, `date measured`,
`recorded sample name`, `processed sample name`,
`Qubit HS RNA (ng/uL)`, `Qubit HS dsDNA (ng/uL)`, `dilution factor`.
We reshape to long format and derive concentration columns that
handle out-of-range readings.

```{r}
qubit_raw <- here("data-raw", "qubit", "measurements.csv") |>
  read_csv(
    col_types = cols(
      handler = "c",
      `date measured` = "D",
      `recorded sample name` = "c",
      `processed sample name` = "c",
      `Qubit HS RNA (ng/uL)` = "c",
      `Qubit HS dsDNA (ng/uL)` = "c",
      `dilution factor` = "d"
    ),
    na = ""
  )
```

```{r}
# Assay detection limits used to bound out-of-range readings.
rna_lower <- 4; rna_upper <- 200
dna_lower <- 0.1; dna_upper <- 120

qubit <- qubit_raw |>
  pivot_longer(
    cols = matches("^Qubit.*\\(ng/uL\\)$"),
    names_to = "assay",
    values_to = "conc_chr"
  ) |>
  janitor::clean_names() |>
  rename(
    measurement_date = date_measured,
    processed_sample_id = processed_sample_name,
    recorded_sample_id = recorded_sample_name
  ) |>
  filter(conc_chr != "none", startsWith(processed_sample_id, "PR-")) |>
  relocate(processed_sample_id) |>
  mutate(
    na_type = str_extract(assay, "(?<=^Qubit HS )(dsDNA|RNA)"),
    measurement_year = year(measurement_date) |> as.factor(),
    is_low_oor = conc_chr == "Low OOR",
    is_high_oor = conc_chr == "High OOR",
    conc_machine = case_when(
      conc_chr %in% c("Low OOR", "High OOR") ~ NA_real_,
      TRUE ~ as.numeric(conc_chr)
    ),
    range_status = case_when(
      conc_chr %in% c("Low OOR", "High OOR") ~ conc_chr,
      TRUE ~ "In range"
    ),
    input_volume = case_when(
      measurement_date == as.Date("2024-02-22") & assay == "Qubit HS RNA (ng/uL)" ~ 5,
      TRUE ~ 2
    ),
    conc_machine_oor = case_when(
      na_type == "RNA"   & conc_chr == "Low OOR"  ~ rna_lower / input_volume,
      na_type == "RNA"   & conc_chr == "High OOR" ~ rna_upper / input_volume,
      na_type == "dsDNA" & conc_chr == "Low OOR"  ~ dna_lower / input_volume,
      na_type == "dsDNA" & conc_chr == "High OOR" ~ dna_upper / input_volume,
      TRUE ~ conc_machine
    ),
    conc_sample = conc_machine * dilution_factor,
    conc_sample_oor = conc_machine_oor * dilution_factor
  )

write_rds(qubit, here("data", "qubit.rds"))
```

## Quick sanity checks

```{r}
qpcr_results |> count(target, task)
qubit |> count(assay, range_status)
```
