---
title: "Prepare MGS data for the manuscript"
author: "Mike"
date: today
format:
  html:
    toc: true
    fig-format: svg
    fig-dpi: 300
    fig-width: 8
    fig-asp: 0.618
---

This script reads the raw outputs of the NAO Viral MGS Analysis Pipeline
under `data-raw/mgs/` and builds two `phyloseq` objects in `data/`:

- `data/kraken-phyloseq.rds` — Kraken2 clade abundances. Each feature
  is a `(taxid, ribosomal)` pair so ribosomal vs. non-ribosomal reads
  are kept as separate features.
- `data/hv-phyloseq.rds` — Human-virus clade counts from the
  pipeline's HV workflow.

Both objects carry sample metadata via the BMC library manifest
(joined to the processed-sample log).

The Kraken2 standard-DB inspect file is large (~3 GB) and is hosted
publicly by the Kraken2 maintainers; this script downloads it on first
run if `data-raw/public/kraken-standard_20240605-inspect.txt` is
missing. See `data-raw/public/DOWNLOAD.md` for the source URL.

```{r setup}
#| output: false
library(tidyverse)
library(fs)
library(here)
library(speedyseq)
options(warn = 1)  # print warnings as they occur

source(here("code", "_helpers.R"))
```

## Sample metadata

```{r}
meta_processed <- here("data", "processed-samples-metadata.rds") |>
  read_rds()

meta_libraries <- here("data", "libraries.tsv") |>
  read_tsv(show_col_types = FALSE)
```

## Pipeline outputs

```{r}
mgs_path <- here("data-raw", "mgs")

stats <- path(mgs_path, "results", "qc", "qc_basic_stats.tsv.gz") |>
  read_tsv(show_col_types = FALSE) |>
  mutate(stage = str_replace(stage, "raw_concat", "raw")) |>
  arrange(sample)

stats_wide <- stats |>
  select(sample, stage, n_read_pairs) |>
  pivot_wider(
    names_from = stage, values_from = n_read_pairs,
    names_prefix = "n_read_pairs_"
  )
```

## Build per-(submission-sample) metadata

The pipeline combines libraries with the same submission_sample_id, so
metadata at the analysis level is keyed by `submission_sample_id` (here
renamed to `sample`).

```{r}
meta <- meta_libraries |>
  select(-library_id, -index_sequences, -provider_prep_batch) |>
  distinct() |>
  mutate(.after = library_preparation_method,
    ribo_depleted = str_detect(library_preparation_method, "rRNA Depletion")
  ) |>
  mutate(
    sample_type = fct_relevel(sample_type, "sludge", "influent", "airplane")
  ) |>
  rename(sample = submission_sample_id) |>
  left_join(stats_wide, by = "sample")
```

## Taxonomy workflow

```{r}
kraken <- path(mgs_path, "results", "taxonomy", "kraken_reports_merged.tsv.gz") |>
  read_tsv(col_types = "nnnnnciccl")
```

Get the Kraken2 standard-DB taxonomy via `kraken2-inspect` output. Cache
the subset to `_output/` so this slow step runs once per repo clone.

```{r}
inspect_path <- here("data-raw", "public", "kraken-standard_20240605-inspect.txt")
if (!file.exists(inspect_path)) {
  message("Downloading kraken2-inspect output (~3 GB)...")
  download.file(
    "https://genome-idx.s3.amazonaws.com/kraken/standard_20240605/inspect.txt",
    inspect_path
  )
}

cache_path <- here("_output", "kraken_inspect_subset.rds")
if (!file.exists(cache_path)) {
  message("Parsing inspect.txt (slow; cached after first run)...")
  kraken_inspect <- inspect_path |>
    read_kraken_reports(minimizers = FALSE) |>
    select(-file) |>
    kraken_add_taxonomy()
  kraken_inspect_subset <- kraken_inspect |>
    filter(taxid %in% kraken$taxid)
  rm(kraken_inspect)
  dir_create(path_dir(cache_path))
  write_rds(kraken_inspect_subset, cache_path)
} else {
  kraken_inspect_subset <- read_rds(cache_path)
}
```

Build a wide taxonomy table restricted to the standard Linnaean ranks.

```{r}
rns <- c("root", "domain", "kingdom", "phylum", "class",
         "order", "family", "genus", "species")
names(rns) <- rns |> map_chr(\(x) str_sub(x, 1, 1) |> str_to_upper())
rns_map <- set_names(names(rns), rns)

tax <- kraken_inspect_subset |>
  filter(str_detect(rank_code, "[0-9]", negate = TRUE)) |>
  select(taxid, rank_code, taxonomy) |>
  mutate(
    taxonomy = taxonomy |>
      str_remove(";$") |>
      str_split(";") |>
      map(\(x) str_split_fixed(x, "__", 2) |>
        as_tibble(.name_repair = \(x) c("path_rank", "path_name"))
      )
  ) |>
  unnest(taxonomy) |>
  filter(str_detect(path_rank, "[0-9]", negate = TRUE)) |>
  pivot_wider(names_from = path_rank, values_from = path_name) |>
  # Change rank names from single-letter abbreviation to full name; and ensure
  # ranks are in the correct order; needed because Kingdom only appears for
  # eukaryotes and so can end up last
  select(rank = rank_code, any_of(rns_map), taxid) |>
  mutate(rank = fct_recode(rank, !!!rns_map)) |>
  # Add row for unclassified
  add_row(.before = 1, taxid = 0L, rank = "unclassified")
```

### Assemble phyloseq object

```{r}
otu_split <- kraken |>
  arrange(taxid, ribosomal) |>
  transmute(
    .feature = str_glue("{taxid}_{ribosomal}"),
    sample,
    .abundance = n_reads_clade
  ) |>
  pivot_wider(names_from = sample, values_from = .abundance, values_fill = 0) |>
  speedyseq::otu_table(taxa_are_rows = TRUE)

tax_split <- tax |>
  crossing(ribosomal = c(FALSE, TRUE)) |>
  unnest(ribosomal) |>
  mutate(.before = 1, feature = str_glue("{taxid}_{ribosomal}")) |>
  relocate(rank, .after = last_col()) |>
  # We convert taxid to string before passing to tax_table, otherwise tax_table
  # will add whitespace padding to small numbers during its own conversion
  mutate(across(taxid, as.character)) |>
  tax_table()

ps <- phyloseq(meta |> sample_data(), otu_split, tax_split)
rm(otu_split)

# Per-sample total reads = sum of reads at the unclassified + root nodes
# (taxids 0 and 1), summed across ribosomal and non-ribosomal features.
ss <- ps |>
  filter_tax_table(taxid %in% c("0", "1")) |>
  sample_sums()

ps <- ps |>
  mutate_sample_data(
    sample_sum = unname(ss)
  )

write_rds(ps, here("data", "kraken-phyloseq.rds"))
```

## HV workflow

```{r}
hv_clade_counts <- path(mgs_path, "results", "hv", "hv_clade_counts.tsv.gz") |>
  read_tsv(show_col_types = FALSE)

hv_tax <- hv_clade_counts |>
  select(taxid, rank, name) |>
  distinct() |>
  tax_table()

hv_otu <- hv_clade_counts |>
  select(taxid, sample, n_reads_clade) |>
  pivot_wider(names_from = sample, values_from = n_reads_clade, values_fill = 0) |>
  speedyseq::otu_table(taxa_are_rows = TRUE)

hv_ps <- phyloseq(meta |> sample_data(), hv_otu, hv_tax)

write_rds(hv_ps, here("data", "hv-phyloseq.rds"))
```
